---
output: pdf_document
title: Answers to exercises ordered categorical indicator variables
---

# Exercise 6.2

```{r}
gssdat <- read.table("gss2000.dat", header = TRUE)
colMeans(gssdat)
```

Item `word.c` seems to be most difficult, `word.d` seems most easy.

a) 

```{r, message=FALSE, warning=FALSE}
library("lavaan")
gssmod.2PL <- '
  vocab =~ word.a + word.b + word.c + word.d + word.e + word.f + word.g + 
            word.h + word.i + word.j 
'
gssfit.2PL <- cfa(gssmod.2PL, ordered = paste0("word.", letters[1:10]), 
                  data = gssdat)
summary(gssfit.2PL, standardized = TRUE, fit.measures = TRUE)
```


b) The easiest item is `word.d`, most difficult is item `word.c`.

c) The strongest indicator is `word.d`, the weakest indicator is `word.a`.

d)

```{r, warning=FALSE, message=FALSE}
library("ltm")
gssfit.2PL.IRT <- ltm(gssdat ~ z1)
summary(gssfit.2PL.IRT)
```

Most difficult is now `word.a`, easiest is still `word.c`. Best indicator is now `word.d`, weakest indicator is now `word.a`.

e) 

```{r}
gssmod.1PL <- '
  vocab =~ lambda*word.a + lambda*word.b + lambda*word.c + lambda*word.d + 
              lambda*word.e + lambda*word.f + lambda*word.g + lambda*word.h + 
              lambda*word.i + lambda*word.j 
'
gssfit.1PL <- cfa(gssmod.1PL, ordered = paste0("word.", letters[1:10]),
                  data = gssdat)
fitinds <- c("chisq.scaled", "df", "pvalue.scaled", "cfi.scaled",
             "rmsea.scaled", "srmr")
fitMeasures(gssfit.1PL, fitinds)
fitMeasures(gssfit.2PL, fitinds)
lavTestLRT(gssfit.1PL, gssfit.2PL)
```

According to CFI, SRMR, RMSEA and the likelihood-ratio test, the 2PL should be preferred.

```{r}
gssfit.1PL.IRT <- rasch(gssdat)
anova(gssfit.1PL.IRT, gssfit.2PL.IRT)
``` 

According to AIC, BIC and the likelihood-ratio test, the 2PL model should be preferred.