---
output: pdf_document
title: Answers to exercises ordered categorical indicator variables
---

# Exercise 6.1

Item wordings:

1. The woman decides on her own that she does not wish to have a child.
2. The couple agree that they do not wish to have a child.
3. The woman is not married and does not wish to marry the man.
4. The couple cannot afford any more children.

```{r, message = FALSE, warning = FALSE}
library("ltm")
names(Abortion) <- c(paste0("I", 1:4))
```

a) Find the proportion who endorsed each item (i.e., the mean score).

```{r}
colMeans(Abortion)
```

Item `I3` is the most-endorsed (easiest) item, item `I1` is the least endorsed (most difficult) item. 

b) Fit a CFA for binary responses using the CFA function, assuming a single latent variable underlies the item responses.

```{r, warning = FALSE, message = FALSE}
library("lavaan")
model <- '
  lib_ab_views =~ I1 + I2 + I3 + I4
'
fit.abo <- cfa(model, data = Abortion, ordered = paste0("I", 1:4))
summary(fit.abo, standardized = TRUE, fit.measures = TRUE)
```

c) The robust $\chi^2$ value are significant, which is to be expected with a sample size of `r nrow(Abortion)`. The robust CFI indicates good model fit, as does the SRMR. The robust RMSEA indicates that the model does not fit well, and the $p$-value of the close fit test indicates that close fit should be rejected.

Looking at the standardized loadings, all are significant and substantial. All loadings have similar values. The variance of the latent trait is significant.

All in all, I would conclude that model fit seems acceptable.

d) If you would have to create a 1-item abortion attitude test, I would use Item 3, because it has the highest discrimination parameter.

e) If the 1-item test has to be used to find persons with extremely liberal views on abortion, I would select the item with the highest threshold (difficulty): Item 1. Persons agreeing with this statement have relatively the most liberal views on abortion.